BHPWAVE: An adiabatic gravitational waveform model for compact objects undergoing quasi-circular inspirals into rotating massive black holes

### 2023-10-21 ms.tex
Updates from Overleaf

\begin{abstract}
We present \texttt{bhpwave}: a new Python-based, open-source tool for generating the gravitational waveforms of stellar-mass compact objects undergoing quasi-circular inspirals into rotating massive black holes. These binaries, known as extreme-mass-ratio inspirals (EMRIs), are exciting millihertz gravitational wave sources for future space-based detectors. Relativistic models of EMRI gravitational wave signals are necessary to unlock the full scientific potential of these astrophysical sources, yet few open-source EMRI waveform models exist. To fill this void we built \texttt{bhpwave}, which uses the adiabatic approximation from black hole perturbation theory to rapidly construct gravitational waveforms based on the leading-order inspiral dynamics of the binary. Crucially \texttt{bhpwave} is the first open-source adiabatic model to produce waveforms for binaries with \emph{rotating} massive black holes, opening a new dimension of parameter space for astrophysical investigation.
\end{abstract}

@@

We can understand the relative impact of these self-forces on EMRI gravitational waveforms by expanding the phasing of the EMRI gravitational waveform $\Phi_\mathrm{GW}$ in powers of $\epsilon$ via a two-timescale analysis \cite{HindFlan08},
\begin{align} \label{eqn:twotimescale}
\Phi_\mathrm{GW} = \frac{1}{\epsilon} \left[ \Phi_\mathrm{0PA} + \epsilon^{1/2} \Phi_\mathrm{res} + \epsilon \Phi_\mathrm{1PA} + O(\epsilon^2)\right].
\end{align}
The leading-order \emph{adiabatic} term\footnote{This term is also referred to as the post-0 adiabatic or 0PA term to remain consistent with the naming conventions of the sub-leading terms (e.g., $\Phi_\mathrm{1PA}$).} $\Phi_\mathrm{0PA}$ only depends on the time-averaged \emph{dissipative} first-order GSF, which drives the dissipation of energy and angular momentum from the system. The sub-leading post-1 adiabatic (1PA) piece $\Phi_\mathrm{1PA}$ depends on the remaining contributions from the first-order GSF and the time-averaged components of the second-order GSF. The half-order correction $\Phi_\mathrm{res}$ arises due to the presence of self-forced $r\theta$-resonances experienced by EMRIs undergoing eccentric and inclined inspirals \cite{FlanHind12}. EMRI gravitational models must, therefore, include the $\Phi_\mathrm{0PA}$, $\Phi_\mathrm{res}$, and $\Phi_\mathrm{1PA}$ effects in order to maintain sub-radian phase accuracy and enable the full scientific potential of future mHz gravitational wave observatories.

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (but not characterizing) EMRI signals and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing EMRIs \cite{HughETC21}.

In practice, however, it is challenging to optimize adiabatic EMRI waveform calculations in order to make them efficient and accessible, so that they can be incorporated in large scale samplings of parameter space. Currently, only one open-source tool exists for producing adiabatic EMRI waveforms: the Fast EMRI Waveforms (FEW) Python package \cite{ChuaGallVall19, ChuaETC20, KatzETC20, KatzETC21}. While this tool is the gold standard for EMRI waveform models, particularly due to its ability to leverage GPUs, it is currently restricted to eccentric binaries with non-rotating black holes. Since we expect almost all astrophysical EMRIs to possess a rotating massive black hole, this model misses a crucial area of parameter space.

Therefore we introduce \texttt{bhpwave} [RTD link?] an open-source Python-based waveform generator that models the adiabatic dynamics and gravitational wave signals produced by a small body undergoing a quasi-circular\footnote{By ``quasi-circular," we mean that, at any moment of time, the inspiral motion is tangent to a circular geodesic as described in Sec.~\ref{sec:adiabatic}. This is a consequence of the fact that in the adiabatic limit, circular orbits remain circular, i.e., $e(t) = O(\epsilon)$ \cite{KennOri96, Kenn98}.} (non-eccentric) inspiral into a \emph{rotating} massive black hole. This code can model any binary with an initial orbital separation of $r_0 \leq 50 M$ and a massive black hole spin in the range $|a| \leq 0.995 M$. (See Sec.~\ref{sec:adiabatic} for exact definitions of $a$ and $r_0$.) Furthermore, the model supports any mass-ratio, though adiabatic waveform models are most relevant for $\epsilon \lesssim 10^{-4}$. By leveraging parallel computations across CPUs, \texttt{bhpwave} can evaluate years-long waveforms in milliseconds. Even on a standard laptop, waveform evaluations still complete within a few hundred milliseconds to a couple seconds.

@@

Additionally, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, reliable, and accessible. Even when neglecting eccentricity, the inclusion of massive black hole spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher mode content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in this area of parameter space and provides an important stepping stone for developing more advanced models that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors}, we use a toy problem test the impact of modeling errors on parameter biases for EMRI data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

The mass $\mu$ follows a geodesic $z_p^\mu \doteq (t_p, r_p, \theta_p, \phi_p)$ that maintains a constant Boyer-Lindquist radius $r_p = r_0$ and is restricted to the equatorial plane $\theta_p = \pi/2$ (with respect to the angular momentum of the Kerr black hole). Due to the Killing symmetries of Kerr spacetime, this motion possesses three constants of motion: the orbital energy $u_t = - \mathcal{E}$, the $z$-component of the orbital angular momentum $u_\phi = \mathcal{L}_z$, and the Carter constant $\mathcal{Q} = Q_{\mu\nu} u^\mu u^\nu$ [citation], which are related to $a$ and $r_0$ by
\begin{subequations} \label{eqn:En}
\begin{align}
    \mathcal{E} &= \frac{1 - 2 v^2 \pm \hat{a} v^3}{\sqrt{1-3v^2\pm 2\hat{a}v^3}},
    \\
    \mathcal{L}_z &= \pm M  \frac{1 \mp 2 \hat{a} v^3 + \hat{a}^2 v^4}{v\sqrt{1-3v^2\pm 2\hat{a}v^3}},
    \\
    \mathcal{Q} &= 0,
\end{align}
\end{subequations}
where $\hat{a} \equiv a/M$, $v^2 = M/r_0$, and $+$ ($-$) refers to prograde (retrograde) orbits.

@@

The spin-weighted spheroidal harmonics can be conveniently represented as a rapidly-converging series of spin-weighted spherical harmonics $Y_{s\ell m}(\theta)e^{im\phi}$ \cite{Hugh00b},
\begin{align}
    S_{sjm}(\theta, \gamma) = \sum_{\ell=\ell_\mathrm{min}}^\infty b_{sjm}^\ell(\gamma) Y_{s\ell m}(\theta),
\end{align}
where $\ell_\mathrm{min}=\mathrm{max}[|m|,|s|]$. This decomposition is particularly useful because it allows us to reproject $\psi_4$ onto an angular basis that is independent of frequency,
\begin{align} \label{eqn:psi4LModes}
    \psi_4 = \rho^{4} \sum_{\ell=2}^\infty \sum_{m=-\ell}^\ell X_{-2\ell m}(r) Y_{-2\ell m}(\theta)e^{im(\phi - \Omega_pt)},
\end{align}
where
\begin{align} \label{eqn:XfromR}
    X_{-2\ell m}(r) = \sum_{j=\ell_\mathrm{min}}^\infty b_{-2jm}^\ell(\gamma)R_{-2jm}(r).
\end{align}
Furthermore, the spin-weighted spherical harmonics form a complete and orthonormal set of basis functions on the unit sphere (for the same value of spin-weight),
\begin{align}
    \int Y_{s\ell m} {Y}_{s\ell'm'} e^{i(m-m')\phi} d\Omega = \delta_{l l'} \delta_{m m'}.
\end{align}

@@

For our point-particle source on a circular geodesic, the radial mode function $R_{-2jm}(r)$ have the asymptotic forms
\begin{subequations} \label{eqn:asympR}
\begin{align}
    R_{-2jm}(r\rightarrow r_+) &\simeq Z^\mathcal{H}_{-2jm} \Delta^{2} e^{-im k_p r_*},
    \\ 
    R_{-2jm}(r\rightarrow \infty) &\simeq Z^\mathcal{I}_{-2jm} r^{3} e^{+im\Omega_p r_*},
\end{align}
\end{subequations}
where $k_p = \Omega_p - \Omega_+$, the horizon frequency $\Omega_+ = a/(2Mr_+)$, and the tortoise coordinate $r_*$ is given by the differential relation $dr_*/dr = (r^2+a^2)/\Delta$.
The amplitudes $Z^\mathcal{H/I}_{-2jm}$ are often referred to as \emph{Teukolsky amplitudes} and are constructed via the standard Green's function method, also known as the method of variation of parameters (see Sec.~III\,A in \cite{HughETC21} or \cite{DrasHugh06} for further details). From \eqref{eqn:XfromR}, we see that $X_{-2\ell m}$ then possesses the same asymptotic behavior as $R_{-2\ell m}$ in \eqref{eqn:asympR} but with modified amplitudes
\begin{align}
    Z^\mathcal{H/I}_{-2jm} \rightarrow X^\mathcal{H/I}_{-2\ell m} = \sum_{j=\ell_\mathrm{min}}^\infty b^\ell_{-2jm}(\gamma) Z^\mathcal{H/I}_{-2jm}.
\end{align}
As a final note, because $Z^\mathrm{H/I}_{-2jm}$ and  $X^\mathrm{H/I}_{-2\ell m}$ depend on the source motion, we can parametrize both amplitudes in terms of the orbital constants, e.g., $X^\mathcal{H/I}_{-2\ell m}= {X}^\mathcal{H/I}_{-2\ell m}(\Omega_p; a)$.

@@

After determining the adiabatic inspiral of the smaller body, we generalize \eqref{eqn:psi4LModes}---our geodesic expression for $\psi_4$---by once again leveraging the fact that at any moment of time the motion is approximately tangent to a geodesic. Consequently, the field amplitudes are promoted from constants to quantities that slowly evolve with frequency and time $X^\mathcal{H/I}_{-2\ell m}(\Omega_p; a) \rightarrow X^\mathcal{H/I}_{-2\ell m}(\Omega(t); a)$, while the field's phase rapidly accumulates in proportion to the orbital phase $\Omega_p t \rightarrow \Phi(t)$. As a result, the adiabatic waveform $h = h_+ - i h_\times$ takes the form
\begin{subequations} \label{eqn:adiabaticWaveform}
\begin{align} 
    h(u,r,\theta,\phi) &= \frac{\mu}{r}\sum_{\ell m} h_{\ell m}(u,r,\theta,\phi),
    \\
    h_{\ell m}(u,r,\theta,\phi) & = B_{\ell m}(u) Y_{-2\ell m}(\theta) e^{im[\phi - \Phi(u)]}  
\end{align}
\end{subequations}
where, assuming $r \gg M$, the waveform amplitudes are given by
\begin{align} \label{eqn:amplitude}
    B_{\ell m}(u) = -2\frac{X_{-2\ell m}^\mathcal{I}( \Omega(u);a)}{m^2\Omega^2(u)} = A_{\ell m} e^{i \psi_{\ell m}}.
\end{align}
For later convenience we introduce the magnitude of the waveform amplitudes $A_{\ell m} = |B_{\ell m}|$ and the phase of the complex amplitudes $\psi_{\ell m}$, both of which depend on $\Omega(u)$ and $a$. Furthermore, rather than using the conventions of \cite{KatzETC21,HughETC21}, we follow \cite{PounWard20} and parametrize the adiabatic waveform in terms of the outgoing time-coordinate $u=t-r_*$. As a result, if we hold the orbital constants and frequencies fixed, \eqref{eqn:adiabaticWaveform} reduces to a geodesic ``snapshot'' waveform obtained via \eqref{eqn:psi4ToH} and \eqref{eqn:psi4LModes}. In practice, when implementing \eqref{eqn:adiabaticWaveform} in \texttt{bhpwave}, we replace $u$ with Boyer-Lindquist time $t$ to match the FEW model (which is described by (10) in \cite{KatzETC21}). This is equivalent to parametrizing all of our waveforms with an initial time $t_0 = r_*$.

@@

\section{Solar system barycenter waveforms}

Up to this point, waveforms have been constructed in the source frame, with our Boyer-Lindquist coordinate system centered on the massive black hole. To get the observed waveform in the solar system barycenter (SSB) frame $h_\mathrm{SSB}$, we adopt the same conventions as the \textsc{FEW} model and make use of the transformation provided in \cite{KatzETC21}. In this frame, 
a generic EMRI system is parametrized by 12 intrinsic parameters $(\mu, M, a, \vec{a}_2, p_0, e_0, x_0, \Phi_{r0}, \Phi_{\theta 0}, \Phi_{\phi 0})$---where $\vec{a}_2$ is the spin-vector of the smaller body; $p_0$, $e_0$, and $x_0$ are respectively the initial semi-latus rectum, orbital eccentricity, and projection of the orbital inclination; and $\Phi_{r0}$, $\Phi_{\theta 0}$, $\Phi_{\phi 0}$ are respectively the initial radial, polar, and azimuthal phases---and by 5 extrinsic parameters $(d_L, q_S, \phi_S, q_K, \phi_K)$---where $d_L$ is the luminosity distance to the source EMRI, $q_S$ and $\phi_S$ are the polar and azimuthal sky positions of the source, and $q_K$ and $\phi_K$ are the polar and azimuthal angles defining the orientation of the massive black hole's axis of rotation. In our simplified quasi-circular setup, five of the intrinsic parameters are constrained to the values $\vec{a}_2 = \vec{0}$, $e_0 = 0$, $x_0 = 1$, $\Phi_{r0} = \Phi_{\theta 0} = 0$, while two of the remaining free intrinsic parameters are given by $\Phi_{\phi 0} = \Phi(0)$ and $p_0 = r_0$.

The extrinsic parameters are related to Boyer-Lindquist $(r,\theta, \phi)$ via
\begin{subequations} \label{eqn:SSBtoSource}
    \begin{align}
        r &= d_L,
        \\
        \cos\theta &= - \cos q_S\cos q_K
        \\ \notag
        & \qquad \quad -\sin q_S \sin q_K \cos(\phi_K - \phi_S),
        \\
        \phi &= -\frac{\pi}{2}.
    \end{align}
\end{subequations}
The frame transformation also rotates the polarization basis by the polarization angle $\psi$ \cite{KatzETC21}. As a result, $h_\mathrm{SSB}$ is related to $h$, the strain in the source frame, via
\begin{align} \label{eqn:polarization}
    h_\mathrm{SSB} = e^{2i\psi} h = \frac{D_\psi^2}{|D_\psi|^2} h = \frac{D_\psi}{D^*_\psi} h,
\end{align}
where $D_\psi$ is given by
\begin{multline}
    D_\psi = \cos q_S \sin q_K \cos(\phi_K - \phi_S)
    \\ 
     - \sin q_S\cos q_K +i \sin q_K \sin(\phi_K - \phi_S),
\end{multline}
and $a^*$ denotes the complex conjugate of $a$. Note that there are two cases in which the preferred wave basis in \cite{KatzETC20} is no longer uniquely defined: $(q_K, \phi_K) = (q_S, \phi_S)$ and $(q_K, \phi_K) = (\pi - q_S, \pi + \phi_S)$. In both instances the SSB and source frames are aligned or anti-aligned with one another, leading to $D_\psi = 0$ and invalidating the last equality in \eqref{eqn:polarization}. Therefore, we choose $h_\mathrm{SSB} = h$ if $(q_K, \phi_K) = (q_S, \phi_S)$ or $(q_K, \phi_K) = (\pi - q_S, \pi + \phi_S)$.

@@

\subsection{Equations of motion and the numerical domain}

@@

\subsection{Interpolated fluxes and trajectories}

>>>>

\begin{abstract}
We present \texttt{bhpwave}: a new Python-based, open-source tool for generating the gravitational waveforms of stellar-mass compact objects undergoing quasi-circular inspirals into rotating massive black holes. These binaries, known as extreme-mass-ratio inspirals (EMRIs), are exciting millihertz gravitational wave sources for future space-based detectors. Relativistic models of EMRI gravitational wave signals are necessary to unlock the full scientific potential of these astrophysical sources, yet few open-source EMRI waveform models exist. To fill this void we built \texttt{bhpwave}, which uses the adiabatic approximation from black hole perturbation theory to rapidly construct gravitational waveforms based on the leading-order inspiral dynamics of the binary. Crucially \texttt{bhpwave} is the first open-source adiabatic model to produce waveforms for binaries with \emph{rotating} massive black holes, opening a new dimension of parameter space for astrophysical investigation. In this work, we present the theoretical and numerical foundations underpinning \texttt{bhpwave}. We also use \texttt{bhpwave} to perform a couple basic assessments of how EMRI modeling errors can impact LISA gravitational wave data analysis. 
\end{abstract}

@@

We can understand the relative impact of these self-forces on EMRI gravitational waveforms by expanding the phasing of the EMRI gravitational waveform $\Phi_\mathrm{GW}$ in powers of $\epsilon$ via a two-timescale analysis \cite{HindFlan08},
\begin{align} \label{eqn:twotimescale}
\Phi_\mathrm{GW} = \frac{1}{\epsilon} \left[ \Phi_\mathrm{0PA} + \epsilon^{1/2} \Phi_\mathrm{res} + \epsilon \Phi_\mathrm{1PA} + O(\epsilon^2)\right].
\end{align}
The leading-order \emph{adiabatic} term\footnote{This term is also referred to as the post-0 adiabatic or 0PA term to remain consistent with the naming conventions of the sub-leading terms (e.g., $\Phi_\mathrm{1PA}$).} $\Phi_\mathrm{0PA}$ only depends on the time-averaged \emph{dissipative} first-order GSF, which drives the dissipation of energy and angular momentum from the system. The sub-leading post-1 adiabatic (1PA) piece $\Phi_\mathrm{1PA}$ depends on the remaining contributions from the first-order GSF, corrections due to the spin of the smaller body, and the time-averaged components of the second-order GSF. The half-order correction $\Phi_\mathrm{res}$ arises due to the presence of self-forced $r\theta$-resonances experienced by EMRIs undergoing eccentric and inclined inspirals \cite{FlanHind12}. EMRI gravitational models must, therefore, include the $\Phi_\mathrm{0PA}$, $\Phi_\mathrm{res}$, and $\Phi_\mathrm{1PA}$ effects in order to maintain sub-radian phase accuracy and enable the full scientific potential of future mHz gravitational wave observatories.

Nonetheless purely adiabatic models---which only include $\Phi_\mathrm{0PA}$ in the gravitational wave phasing---may be suitable for detecting (and possibly characterizing) EMRI signals and are invaluable tools for developing data analysis pipelines for EMRI search and parametrization. They capture almost all of the phasing information and relativistic behavior of EMRI signals, making them powerful probes of astrophysical parameter space. The theoretical foundation and numerical calculation of adiabatic waveforms is also well understood, and has been performed for eccentric, precessing EMRIs \cite{HughETC21}.

However, in practice it is challenging to optimize adiabatic EMRI waveform calculations in order to make them efficient and accessible, so that they can be incorporated in large scale samplings of parameter space. Currently, only one open-source tool exists for producing adiabatic EMRI waveforms: the Fast EMRI Waveforms (FEW) Python package \cite{ChuaGallVall19, ChuaETC20, KatzETC20, KatzETC21}. While this tool is the gold standard for EMRI waveform models, particularly due to its ability to leverage GPUs, it is currently restricted to eccentric binaries with non-rotating black holes. Since we expect almost all astrophysical EMRIs to possess a rotating massive black hole, it is important to extend models such as FEW into this crucial area of parameter space.

Therefore we introduce \texttt{bhpwave} [RTD link?] an open-source Python-based waveform generator that models the adiabatic dynamics and gravitational wave signals produced by a small body undergoing a quasi-circular\footnote{By ``quasi-circular," we mean that, at any moment of time, the inspiral motion is tangent to a circular geodesic as described in Sec.~\ref{sec:adiabatic}. This is a consequence of the fact that in the adiabatic limit, circular orbits remain circular, i.e., $e(t) = O(\epsilon)$ \cite{KennOri96, Kenn98}.} (non-eccentric) inspiral into a {rotating} massive black hole. This code can model any binary with an initial orbital separation of $r_0 \leq 60 M$ and a massive black hole spin in the range $|a| \leq 0.9999 M$. (See Sec.~\ref{sec:adiabatic} for exact definitions of $a$ and $r_0$.) Furthermore, the model supports any mass-ratio, though adiabatic waveform models are most relevant for $\epsilon \lesssim 10^{-4}$. By leveraging parallel computations across CPUs, \texttt{bhpwave} can evaluate years-long waveforms in milliseconds. Even on a standard laptop, waveform evaluations still complete within a few hundred milliseconds to a couple seconds.

@@

Additionally, \texttt{bhpwave} does not provide the first adiabatic calculation of quasi-circular inspirals. These systems have long been studied in the literature \cite{Detw78, KennOri96, FinnThor00, Hugh00b, GralHughWarb16}. However, as aforementioned, it is challenging to implement an adiabatic model that is efficient, reliable, and accessible. Even when neglecting eccentricity, the inclusion of massive black hole spin presents a number of challenges that are less pronounced for binaries with non-spinning bodies, such as the larger range of radial separations accessible to spinning binaries, the faster frequency evolution for more deeply bound orbits, and the higher harmonic content needed for modeling rapidly-rotating systems. By building upon the theoretical foundations described in the literature, our aim for \texttt{bhpwave} is to provide an efficient and accessible model that fills in a new area of parameter space and provides an important stepping stone for developing more advanced models that incorporate all possible orbital effects.

To encourage the continued development of open-source adiabatic EMRI models, this paper outlines the theoretical and numerical foundations underpinning \texttt{bhpwave}. In Sec.~\ref{sec:adiabatic} we summarize the quasi-circular limit of the adiabatic approximation within the context of black hole perturbation theory to establish methods and notation. 
%In Sec.~\ref{sec:bhpwave} we outline how \texttt{bhpwave} implements the quasi-circular adiabatic model, primarily through its three modules: \texttt{bhpwave.trajectory}, \texttt{bhpwave.harmonics}, and \texttt{bhpwave.waveform}. 
In Sec.~\ref{sec:numerical} we describe both the numerical routines used to generate the adiabatic data for \texttt{bhpwave} and the algorithms employed by \texttt{bhpwave} to generate quasi-circular inspirals, waveform harmonics, and gravitational signals. Additionally, we provide validation tests and model comparisons to verify the accuracy of \texttt{bhpwave}. To demonstrate the utility of \texttt{bhpwave}, in Sec.~\ref{sec:errors} we use a few example problems test the impact of modeling errors on parameter biases for EMRI data analysis. Finally, we discuss further applications and possible extensions of \texttt{bhpwave} in Sec.~\ref{sec:conclusion}. For this paper we use the metric signature $(-+++)$, the sign conventions, where applicable, of \cite{MisnThorWhee73}, and units such that $G=c=1$.

@@

The mass $\mu$ follows a geodesic $z_p^\mu \doteq (t_p, r_p, \theta_p, \phi_p)$ that maintains a constant Boyer-Lindquist radius $r_p = r_0$ and is restricted to the equatorial plane $\theta_p = \pi/2$ (with respect to the angular momentum of the Kerr black hole). Due to the Killing symmetries of Kerr spacetime, this motion possesses three constants of motion: the orbital energy $u_t = - \mathcal{E}$, the $z$-component of the orbital angular momentum $u_\phi = \mathcal{L}_z$, and the Carter constant $\mathcal{Q} = Q_{\mu\nu} u^\mu u^\nu$ \cite{SteiWarb}, which are related to $a$ and $r_0$ by
\begin{subequations} \label{eqn:En}
\begin{align}
    \mathcal{E} &= \frac{1 - 2 v^2 \pm \hat{a} v^3}{\sqrt{1-3v^2\pm 2\hat{a}v^3}},
    \\
    \mathcal{L}_z &= \pm M  \frac{1 \mp 2 \hat{a} v^3 + \hat{a}^2 v^4}{v\sqrt{1-3v^2\pm 2\hat{a}v^3}},
    \\
    \mathcal{Q} &= 0,
\end{align}
\end{subequations}
where $\hat{a} \equiv a/M$, $v^2 = M/r_0$, and $+$ ($-$) refers to prograde (retrograde) orbits.

@@

The spin-weighted spheroidal harmonics can be conveniently represented as a rapidly-converging series of spin-weighted spherical harmonics $Y_{s\ell m}(\theta)e^{im\phi}$ \cite{Hugh00b},
\begin{align}
    S_{sjm}(\theta, \gamma) = \sum_{\ell=\ell_\mathrm{min}}^\infty b_{sjm}^\ell(\gamma) Y_{s\ell m}(\theta),
\end{align}
where $\ell_\mathrm{min}=\mathrm{max}[|m|,|s|]$. This decomposition is particularly useful because we can then reproject $\psi_4$ onto an angular basis that is independent of frequency,
\begin{align} \label{eqn:psi4LModes}
    \psi_4 = \rho^{4} \sum_{\ell=2}^\infty \sum_{m=-\ell}^\ell X_{-2\ell m}(r) Y_{-2\ell m}(\theta)e^{im(\phi - \Omega_pt)},
\end{align}
where
\begin{align} \label{eqn:XfromR}
    X_{-2\ell m}(r) = \sum_{j=\ell_\mathrm{min}}^\infty b_{-2jm}^\ell(\gamma)R_{-2jm}(r).
\end{align}
Furthermore, the spin-weighted spherical harmonics form a complete and orthonormal set of basis functions on the unit sphere (for the same value of spin-weight),
\begin{align}
    \int Y_{s\ell m} {Y}_{s\ell'm'} e^{i(m-m')\phi} d\Omega = \delta_{l l'} \delta_{m m'}.
\end{align}

@@

For our point-particle source on a circular geodesic, the radial mode function $R_{-2jm}(r)$ have the asymptotic form
\begin{subequations} \label{eqn:asympR}
\begin{align}
    R_{-2jm}(r\rightarrow r_+) &\simeq Z^\mathcal{H}_{-2jm} \Delta^{2} e^{-im k_p r_*},
    \\ 
    R_{-2jm}(r\rightarrow \infty) &\simeq Z^\mathcal{I}_{-2jm} r^{3} e^{+im\Omega_p r_*},
\end{align}
\end{subequations}
where $k_p = \Omega_p - \Omega_+$, the horizon frequency $\Omega_+ = a/(2Mr_+)$, and the tortoise coordinate $r_*$ is given by the differential relation $dr_*/dr = (r^2+a^2)/\Delta$.
The amplitudes $Z^\mathcal{H/I}_{-2jm}$ are often referred to as \emph{Teukolsky amplitudes} and are constructed via the standard Green's function method, also known as the method of variation of parameters (see Sec.~III\,A in \cite{HughETC21} or \cite{DrasHugh06} for further details). From \eqref{eqn:XfromR}, we see that $X_{-2\ell m}$ then possesses the same asymptotic behavior as $R_{-2\ell m}$ in \eqref{eqn:asympR} but with modified amplitudes
\begin{align}
    Z^\mathcal{H/I}_{-2jm} \rightarrow X^\mathcal{H/I}_{-2\ell m} = \sum_{j=\ell_\mathrm{min}}^\infty b^\ell_{-2jm}(\gamma) Z^\mathcal{H/I}_{-2jm}.
\end{align}
As a final note, because $Z^\mathrm{H/I}_{-2jm}$ and  $X^\mathrm{H/I}_{-2\ell m}$ depend on the source motion, we can parametrize both amplitudes in terms of the orbital constants, e.g., $X^\mathcal{H/I}_{-2\ell m}= {X}^\mathcal{H/I}_{-2\ell m}(\Omega_p; a)$.

@@

After determining the adiabatic inspiral of the smaller body, we generalize \eqref{eqn:psi4LModes}---our geodesic expression for $\psi_4$---by once again leveraging the fact that at any moment of time the motion is approximately tangent to a geodesic. Consequently, the field amplitudes are promoted from constants to quantities that slowly evolve with frequency and time $X^\mathcal{H/I}_{-2\ell m}(\Omega_p; a) \rightarrow X^\mathcal{H/I}_{-2\ell m}(\Omega(t); a)$, while the field's phase rapidly accumulates in proportion to the orbital phase $\Omega_p t \rightarrow \Phi(t)$. As a result, the adiabatic waveform $h = h_+ - i h_\times$ takes the form
\begin{subequations} \label{eqn:adiabaticWaveform}
\begin{align} 
    h(u,r,\theta,\phi) &= \frac{\mu}{r}\sum_{\ell m} h_{\ell m}(u,r,\theta,\phi),
    \\
    h_{\ell m}(u,r,\theta,\phi) & = B_{\ell m}(u) Y_{-2\ell m}(\theta) e^{im[\phi - \Phi(u)]}  
\end{align}
\end{subequations}
where, assuming $r \gg M$, the waveform amplitudes are given by
\begin{align} \label{eqn:amplitude}
    B_{\ell m}(u) = -2\frac{X_{-2\ell m}^\mathcal{I}( \Omega(u);a)}{m^2\Omega^2(u)} = A_{\ell m} e^{i \psi_{\ell m}}.
\end{align}
For later convenience we introduce the magnitude of the waveform amplitudes $A_{\ell m} = |B_{\ell m}|$ and the phase of the complex amplitudes $\psi_{\ell m}$, both of which depend on $\Omega(u)$ and $a$. Furthermore, rather than using the conventions of \cite{KatzETC21,HughETC21}, we follow \cite{PounWard20} and parametrize the adiabatic waveform in terms of the outgoing time-coordinate $u=t-r_*$. As a result, if we hold the orbital constants and frequencies fixed, \eqref{eqn:adiabaticWaveform} reduces to a geodesic ``snapshot'' waveform obtained via \eqref{eqn:psi4ToH} and \eqref{eqn:psi4LModes}. In practice, when implementing \eqref{eqn:adiabaticWaveform} in \texttt{bhpwave}, we replace $u$ with Boyer-Lindquist time $t$ to match the FEW model (which is described by (10) in \cite{KatzETC21}). This is equivalent to parametrizing all of our waveforms with an initial time $t_0 = r_*$. Therefore, we replace $u$ with $t$ for the remainder of this work.

@@

\subsection{Solar system barycenter waveforms}

Up to this point, waveforms have been constructed in the source frame, with our Boyer-Lindquist coordinate system centered on the massive black hole. To get the observed waveform in the solar system barycenter (SSB) frame $h_\mathrm{SSB}$, we adopt the same conventions as the \textsc{FEW} model and make use of the transformation provided in \cite{KatzETC21}. In this frame, 
a generic EMRI system is parametrized by 12 intrinsic parameters $(M, \mu, a, \vec{a}_2, p_0, e_0, x_0, \Phi_{r0}, \Phi_{\theta 0}, \Phi_{\phi 0})$---where $\vec{a}_2$ is the spin-vector of the smaller body; $p_0$, $e_0$, and $x_0$ are respectively the initial semi-latus rectum, orbital eccentricity, and projection of the orbital inclination; and $\Phi_{r0}$, $\Phi_{\theta 0}$, $\Phi_{\phi 0}$ are respectively the initial radial, polar, and azimuthal phases---and by 5 extrinsic parameters $(d_L, q_S, \phi_S, q_K, \phi_K)$---where $d_L$ is the luminosity distance to the source EMRI, $q_S$ and $\phi_S$ are the polar and azimuthal sky positions of the source, and $q_K$ and $\phi_K$ are the polar and azimuthal angles defining the orientation of the massive black hole's axis of rotation. In our simplified quasi-circular setup, five of the intrinsic parameters are constrained to the values $\vec{a}_2 = \vec{0}$, $e_0 = 0$, $|x_0| = 1$, $\Phi_{r0} = \Phi_{\theta 0} = 0$, while two of the remaining free intrinsic parameters are given by $\Phi_{\phi 0} = \Phi(0)$ and $p_0 = r_0$.

The extrinsic parameters are related to Boyer-Lindquist $(r,\theta, \phi)$ via
\begin{subequations} \label{eqn:SSBtoSource}
    \begin{align}
        r &= d_L,
        \\
        \cos\theta &= - \cos q_S\cos q_K
        \\ \notag
        & \qquad \quad -\sin q_S \sin q_K \cos(\phi_K - \phi_S),
        \\
        \phi &= -\frac{\pi}{2}.
    \end{align}
\end{subequations}
The frame transformation also rotates the polarization basis by the polarization angle $\psi$ \cite{KatzETC21}. As a result, $h_\mathrm{SSB}$ is related to $h$, the strain in the source frame, via
\begin{align} \label{eqn:polarization}
    h_\mathrm{SSB} = e^{2i\psi} h = \frac{D_\psi^2}{|D_\psi|^2} h = \frac{D_\psi}{D^*_\psi} h,
\end{align}
where $D_\psi$ is given by
\begin{multline}
    D_\psi = \cos q_S \sin q_K \cos(\phi_K - \phi_S)
    \\ 
     - \sin q_S\cos q_K +i \sin q_K \sin(\phi_K - \phi_S),
\end{multline}
and $a^*$ denotes the complex conjugate of $a$. Note that there are two cases in which the preferred wave basis in \cite{KatzETC21} is no longer uniquely defined: $(q_K, \phi_K) = (q_S, \phi_S)$ and $(q_K, \phi_K) = (\pi - q_S, \pi + \phi_S)$. In these instances the SSB and source frames are aligned or anti-aligned with one another, leading to $D_\psi = 0$ and invalidating the last equality in \eqref{eqn:polarization}. Therefore, we set $h_\mathrm{SSB} = h$ if $(q_K, \phi_K) = (q_S, \phi_S)$ or $(q_K, \phi_K) = (\pi - q_S, \pi + \phi_S)$.

@@

In the following section we discuss the numerical implementation of \texttt{bhpwave}, in particular our process for solving the equations of motion (Secs.~\ref{sec:eom} and \ref{sec:traj}), constructing the harmonic amplitudes of the gravitational wave modes (Sec.~\ref{sec:amp}), and evaluating the total waveform signal (Sec.~\ref{sec:wave}). Finally we compare our model, \texttt{bhpwave}, to the FastEMRIWaveform (FEW) model to test its accuracy. (Sec.~\ref{sec:few}).

\subsection{Equations of motion and the numerical domain}
\label{sec:eom}

@@

\subsection{Interpolated fluxes and trajectories}
\label{sec:traj}
